---
title: "Adding tables into ggplot plots"
author: "karla"
date: '2018-03-19'
slug: adding-tables-into-ggplot-plots
tags:
- rstats
- gridExtra
- dataviz
- ggplot
categories:
- R
- tutorial
---

Sometimes I want to see a table inside a plot in R. 
Let's say that a colleague comes to me with a question: How come there are phenotyping data for a
knockout strain XY even if it's not viable?

In this case I want to see a plot of homozygous/heterozygous pups alongside with a
table of the viability screen outcomes.

Thanks to the `tableGrob` function from the [gridExtra](https://cran.r-project.org/web/packages/gridExtra/vignettes/tableGrob.html) package, this is quite easy.

First, we need to filter and select the data and make the plot.
```{r Grid, warning=FALSE, message=FALSE}
library(tidyverse)
library(gridExtra)

# viability screen outcomes (for the table)
categ <- read_csv('./data/frantisek/categorical_2018-01-19_all.csv')
via <- categ %>% 
  filter(parameter_stable_id == "IMPC_VIA_001_001") %>% 
  filter(gene_symbol == 'Dbn1') %>% 
  select(c("phenotyping_center", "category")) %>% 
  arrange(phenotyping_center) %>% 
  rename(Outcome = category) %>% 
  rename(`Phenotyping center` = phenotyping_center) 

# number of homozygous/heterozygous/wt pups (for plotting)
unidim <- read_csv("./data/frantisek/unidimensional_2018-01-19_all.csv")
pups <- unidim %>% 
  filter(procedure_stable_id == "IMPC_VIA_001") %>% 
  filter(gene_symbol == 'Dbn1') 

cbPalette <- c("#00BA38", "#619CFF", "#F8766D")

plt <- 
pups %>% 
  filter(grepl("^Total pups ", pups$parameter_name)) %>% 
  ggplot(aes(x = phenotyping_center, y = data_point, 
             fill = parameter_name)) +
  geom_bar(stat="identity") +
  scale_fill_manual(values=cbPalette) +
  scale_y_continuous(breaks=seq(0,400,25)) +
  theme_bw() +
  labs(x = "Phenotyping center", y = "Number of pups") +
  theme(legend.title = element_blank(),
        legend.text = element_text(size = 6),
        legend.position = "bottom",
        axis.text = element_text(size = 6, face = "bold"),
        axis.title = element_text(size = 8)) +
  ggtitle("Dbn1 viability") 
```

Now, `gridExtra::tableGrob` creates a table-like plot of a data frame and `annotation_custom()` arranges
it within the plot area.

```{r plot, fig.width= 5, fig.height= 4}
plt + 
   annotation_custom(tableGrob(via, rows=NULL, 
                    theme = ttheme_default(base_size = 5,
                                           padding = unit(c(2, 1), "mm"))), 
                    ymin = 175, xmin = 4)
```


What if I want the table besides the plot? The `grid.arrange` function from the `gridExtra` package will do the trick.

```{r plot2, fig.width=7, fig.height=4}
via_table <- tableGrob(via, rows=NULL, 
                       theme = ttheme_default(base_size = 6))
grid.arrange(plt, via_table, ncol=2, widths = c(2,1))
```

That's it! From this picture, we can see that one of the phenotyping centres find the knockout mice viable and they are able
to produce enough homozygous mice. Don't ask me why.

(All data shown here are publicly available at [IMPC data portal](https://www.mousephenotype.org/data/documentation/data-access).)